---
title: "Session 3: tidyverse and publication-ready figures"
author: "Nils"
date: "`r Sys.Date()`"
output: html_notebook
---

In this session, I will go through the concept of tidy data in R. 
Tidy data is broadly defined as having individual samples in the rows of a data matrix while the columns define variables (or features) describing the samples [https://en.wikipedia.org/wiki/Tidy_data](https://en.wikipedia.org/wiki/Tidy_data).
This is the standard format for any sort of statistical analysis.

We will also discuss the `long` and `wide` data format.

The `tidyverse` library offers a set of libraries that handle tidy data in R:

```{r load-libraries}
library(tidyverse)
```

From the full set of tools, I will be talking about the `readr`, `dplyr`, `tidyr`, `magrittr` and `ggplot2` package in more detail.
This tutorial will closely follow examples provided in the [R for data science book](https://r4ds.had.co.nz/index.html).

The general workflow for data analysis in R is:

1. Reading in your data
2. Tidying your data
3. Transforming your data
4. Visualizing your data
5. Modelling your data

In the last session, we have talked about different ways of reading in the data.

## Introduction

To understand the following session, I will need to introduce a new data type: tibble

### Tibbles

Tibbles are an enhanced version of the base R `data.frame` but with similar functionality.

```{r create-tibble}
data(iris)
iris
as_tibble(iris)
```

Only the first 10 entries are shown.

Tibbles can be created using the `tibble` function:

```{r create-tibble-2}
tb <- tibble(x = 1:5, y = 1, z = x ^ 2 + y)
tb

df <- data.frame(x = 1:5, y = 1)
df$z <- df$x ^ 2 + df$y
df
```

Coercion into `data.frame` works via `as.data.frame`:

```{r coercion}
as.data.frame(tb)
```

Subsetting is enhanced for `tibble` compared to `data.frames` and `stringAsFactor` argument is ignored.

```{r subsetting}
df <- data.frame(abc = 1:3, xyz = c("a", "b", "c"))
df
df$x
df$xyz
df["x"]
df["xyz"]
df[["xyz"]]
df[,"xyz"]
df[,c("abc", "xyz")]

tb <- tibble(abc = 1:3, xyz = c("a", "b", "c"))
tb$x
tb$xyz
tb["x"]
tb["xyz"]
tb[["xyz"]]
tb[,"xyz"]
tb[,c("abc", "xyz")]
```

## Reading in data

We have discussed the base R `read.table` and `read.csv` function in the last session.
Now we will work with the `readr` package to read data into tibbles with a few enhancements:

```{r read_csv}
readr::read_csv("~/Github/IntroDataAnalysis/Data/iris.csv")
```

## The long and wide format

## Data transformation

Wrangling = tidying + transforming

`pivot_longer()`



## Visualize your data

## Model your data

