---
title: "01-readData"
author: "nilseling"
date: "2020-05-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this file, I will read in example data and perform simple exploratory data analysis. 
The toy dataset again is the `iris` datasets that can be loaded via:

```{r load-iris-and-libraries}
library(ggplot2)
data(iris)
```

I usually have to read in much more complex datasets and use the first .Rmd file of my `workflowr` project to read in the data and precess them for subsequent analysis.
Also, I tend to first read in all necessary data and all libraries at the beginning of the script.
Other people prefer reading in data/libraries whenever they are needed.

## Explortory data analysis

As an example, I will generate simple plots to highlight the data and the use of the `workflowr` package.

Here, I will plot `Sepal.Length` vs `Sepal.Width` and colour point by species:

```{r simple-plot}
ggplot(iris) +
    geom_point(aes(x = Sepal.Width, y = Sepal.Length, colour = Species))
```

## Save processed data

In this example, I have not altered the data.
But in more complex cases, I want to save the processed data to read them in in subsequent scripts.
This can be done by storing them as .rds files:

```{r save-data}
saveRDS(object = iris, file = "data/iris.rds")
```



