---
title: "04-plotting"
author: "nilseling"
date: "2020-05-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In todays session, we will go through a number of common plot types and will learn how to generate them using `ggplot2` and other packages of the "ggverse".
For simplicity, I will again use the `iris` dataset and will use this to highlight a couple of different plots.

## Load data and libraries

```{r}
data(iris)
library(ggplot2)
```

As a reminder, the general `ggplot2` call looks like this:

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPING>),
    stat = <STAT>, position = <POSITION>) + 
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION> +
  <THEME_FUNCTION>

## Scatterplot

We will start off with a very simple scatter plot displaying the `Sepal.Length` versus the `Sepal.Width` entry:

```{r scatter-1}
p1 <- ggplot(data = iris) + 
   geom_point(aes(x = Sepal.Length, y = Sepal.Width, colour = Species))
p1
```


With regression lines

Facetting

## Custom PCA plot

## Boxplot

## Barplot

## Heatmap

Using `ggplotify`

## Stitiching it together

Using `cowplot` and `patchwork`.


