---
title: "04-plotting"
author: "nilseling"
date: "2020-05-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In todays session, we will go through a number of common plot types and will learn how to generate them using `ggplot2` and other packages of the "ggverse".
For simplicity, I will again use the `iris` dataset and will use this to highlight a couple of different plots.

## Load data and libraries

```{r}
data(iris)
library(ggplot2)
```

As a reminder, the general `ggplot2` call looks like this:

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPING>),
    stat = <STAT>, position = <POSITION>) + 
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION> +
  <THEME_FUNCTION>

## Scatterplot

We will start off with a very simple scatter plot displaying the `Sepal.Length` versus the `Sepal.Width` entry:

```{r scatter-1}
p1 <- ggplot(data = iris) + 
   geom_point(aes(x = Sepal.Length, y = Sepal.Width, colour = Species))
p1
```

For simple tasks, we can also use the "quickplot" `qplot` function:

```{r qplot}
qplot(x = Sepal.Length, y = Sepal.Width, data = iris, color = Species)
```

### Adding other layers

We can now add other layers to the plot.
Since we can see a linear relationship between Sepal width and Sepal length, we can add some regression lines:

```{r scatter-2}
p1 <- p1 + geom_smooth(aes(x = Sepal.Length, y = Sepal.Width, colour = Species), 
                       method = "lm", formula = "y ~ x")
p1
```

By default, `geom_smooth` will add the 95% confidence interval to the regression line.
By specifying `colour = Species`, a linear regression fit is perform for each level/group of `Species`.

### Facetting

To facilitate visual representation of the three groups, facetting can split the panels based on certain levels:

```{scatter-3}
p1 <- p1 + facet_wrap(. ~ Species)
p1
```

As a reminder, `facet_wrap` is used when splitting by a single variable.
`facet_grid` is used when splitting based on multiple variables.

### Changing the colour scale

Next, we want to change the colour to a more appealing scale.
There are multiple ways to selecting colours.
I like the `RColourBrewer` and `ggsci` packages for discrete colour choice:
Here are some nice resources:

* [https://colorbrewer2.org/](https://colorbrewer2.org/)
* [ggsci](https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html)
* [RColourBrewer](https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette/)

Here, I will use the Nature Publishing Group colour suggestions from the `ggsci` package:

```{r scatter-4}
library(ggsci)
p1 <- p1 + scale_colour_npg()
p1
```

The colours look nicer bu still not perfect.
It is often advisable to store the colour mapping in form of a vector:

```{r scatter-5}
library(RColorBrewer)
display.brewer.all(type = "qual", colorblindFriendly = TRUE)
cur_palette <- brewer.pal(n = 3, name = "Set2")
colour_vec <- c(setosa = cur_palette[1], 
                versicolor = cur_palette[2], 
                virginica = cur_palette[3])
p1 <- p1 + scale_colour_manual(values = colour_vec)
p1
```

We have now changed the colour of the points but are still not quite happy with the plot.

### Changing the theme and aesthetics

First, we might want to use nicer looking points.
For this, we will unfortunatelly need to re-create the whole plot:

```{r scatter-6}
p1 <- ggplot(data = iris) + 
   geom_point(aes(x = Sepal.Length, y = Sepal.Width, fill = Species), 
              shape = 21, size = 2) +
  geom_smooth(aes(x = Sepal.Length, y = Sepal.Width, colour = Species), 
                       method = "lm", formula = "y ~ x") +
  facet_wrap(. ~ Species, 
             labeller = labeller(Species = c("setosa" = "Setosa", # Re-label facet names
                                             "versicolor" = "Versicolor", 
                                             "virginica" = "Virginica"))) +
  scale_fill_manual(values = colour_vec, name = "Iris species",
                    labels = c("Setosa", "Versicolor", "Virginica")) + 
  scale_colour_manual(values = colour_vec, name = "Iris species", 
                      labels = c("Setosa", "Versicolor", "Virginica"))
p1
```

Of note, the "colour" of shape `21` has to be adjusted using the `fill` argument.

Finally, we will change the appearance of the plot:

```{r scatter-7}
p1 <- p1 + xlab("Sepal Width") + ylab("Sepal Length") +
  theme(panel.background = element_blank(), # removes gray colour from the background
        panel.border = element_rect(fill = NA), # Puts a border around the panels
        legend.background = element_rect(colour = "black"), # puts a border around the legend
        legend.position = c(.67, .80), # Re-positions the legend
        axis.line = element_line(), # Adds axis line
        axis.ticks = element_line(colour = "black"), # Re-colours axis ticks
        axis.text = element_text(colour = "black", size = 9), # Re-colours and re-sizes axis tick labels
        axis.title = element_text(colour = "black", size = 12), # Re-colours and re-sizes axis title
        strip.background = element_blank(), # Remove background from strip
        strip.text = element_text(colour = "black", size = 12)) # Re-colours and re-sizes strip text
p1
```

There are nearly infinite adjustable parameters associated to theme: [https://ggplot2.tidyverse.org/reference/theme.html](https://ggplot2.tidyverse.org/reference/theme.html)

I commonly call the entire `ggplot2` function in one call:

```{r create-iris-scatter}
p1 <- ggplot(data = iris) + 
   geom_point(aes(x = Sepal.Length, y = Sepal.Width, fill = Species), 
              shape = 21, size = 2) +
  geom_smooth(aes(x = Sepal.Length, y = Sepal.Width, colour = Species), 
                       method = "lm", formula = "y ~ x") +
  facet_wrap(. ~ Species, 
             labeller = labeller(Species = c("setosa" = "Setosa", 
                                             "versicolor" = "Versicolor", 
                                             "virginica" = "Virginica"))) +
  scale_fill_manual(values = colour_vec, name = "Iris species",
                    labels = c("Setosa", "Versicolor", "Virginica")) + 
  scale_colour_manual(values = colour_vec, name = "Iris species", 
                      labels = c("Setosa", "Versicolor", "Virginica")) +
  xlab("Sepal Width") + ylab("Sepal Length") +
  theme(panel.background = element_blank(), 
        panel.border = element_rect(fill = NA),
        legend.background = element_rect(colour = "black"),
        legend.position = c(.67, .80), 
        axis.line = element_line(),
        axis.ticks = element_line(colour = "black"),
        axis.text = element_text(colour = "black", size = 9),
        axis.title = element_text(colour = "black", size = 12),
        strip.background = element_blank(),
        strip.text = element_text(colour = "black", size = 12))
p1
```

## Custom PCA plot

Two sessions ago, we have created a custom function that plots a certain pca.
We can now use this to display a pca of the iris dataset:

```{r pca-plot}
source("code/auxiliary.R")
cur_pca <- prcomp(iris[,1:4], scale. = TRUE)
p2 <- my_plotPCA(cur_pca, colour_by = iris$Species)
p2
```

Since we coded the function in a way that a `ggplot2` object is returned, we can now make it look prettier:

```{r pca-plot-2}
p2 <- p2 + scale_colour_manual(values = colour_vec, name = "Iris species",
                    labels = c("Setosa", "Versicolor", "Virginica")) +
  coord_fixed() + 
  theme(panel.background = element_blank(),
        legend.background = element_rect(colour = "black"),
        legend.position = c(1, 1), 
        legend.justification = c("right", "top"),
        axis.line = element_line(),
        axis.ticks = element_line(colour = "black"),
        axis.text = element_text(colour = "black", size = 9),
        axis.title = element_text(colour = "black", size = 12))
p2
```

## Boxplot

Next, we will create a simple boxplot to visualize `ggplot2`'s functionality:

```{r boxplot-1}
p3 <- ggplot(iris) + 
  geom_boxplot(aes(x = Species, y = Sepal.Length, fill = Species), 
               outlier.colour = NA, notch = TRUE) + 
  geom_jitter(aes(x = Species, y = Sepal.Length), width = 0.1, alpha = 0.7) + 
  scale_fill_manual(values = colour_vec, name = "Iris species",
                    labels = c("Setosa", "Versicolor", "Virginica")) +
  xlab("") +
  scale_x_discrete(breaks = c("setosa", "versicolor", "virginica"),
                   labels = c("Setosa", "Versicolor", "Virginica")) +
  theme(panel.background = element_blank(),
        legend.background = element_rect(colour = "black"),
        legend.position = c(0.05, 0.95), 
        legend.justification = c("left", "top"),
        axis.line = element_line(),
        axis.ticks = element_line(colour = "black"),
        axis.text.y = element_text(colour = "black", size = 9),
        axis.text.x = element_text(colour = "black", size = 12),
        axis.title = element_text(colour = "black", size = 12))
p3
```

Of note: the "notch" indicates the 95% confidence interval of the median. The two "hindges" represent the 25% and 75% quantiles of the distribution, the upper whisker extends to the value not larger than upper hinge + 1.5 times the interquartile range of the values and the lower whisker extends to the value not smaller than lower hinge - 1.5 times the IQR.

## Barplot

Next, we'll create a barplot visualizing the different measurements for each species.
We also want to add the standard error of the mean to the bar chart - for this we need to modify the data.frame a bit:

```{r pre-process}
library(tidyverse)
cur_dat <- iris %>%
    pivot_longer(cols = colnames(iris)[1:4]) %>%
    group_by(Species, name) %>%
    summarise(mean_val = mean(value),
              sd_val = sd(value),
              n_val = length(value),
              se_val = sd_val/sqrt(n_val),
              ci_t = qt(0.975, df = n_val - 1) * se_val)
```

Now, we can plot the barplot:

```{r barplot}
# Re-order the levels
cur_dat$Measurement <- factor(cur_dat$name, levels = c("Petal.Width", "Petal.Length", "Sepal.Width", "Sepal.Length"))
p4 <- ggplot(cur_dat, aes(x = Species, y = mean_val, fill = Measurement)) + 
  geom_bar(position = position_dodge(),
                           stat = "identity") +
  geom_errorbar(aes(ymin = mean_val - ci_t, ymax = mean_val + ci_t),
                position = position_dodge(0.9), width = 0.2) +
  scale_fill_brewer(palette = "Paired",
                    labels = c("Petal Width", "Petal Length", "Sepal Width", "Sepal Length")) +
  labs(x = "", y = "Measured value [cm]") + 
  scale_x_discrete(breaks = c("setosa", "versicolor", "virginica"),
                   labels = c("Setosa", "Versicolor", "Virginica")) +
  theme(panel.background = element_blank(),
        legend.background = element_rect(colour = "black"),
        legend.position = c(0.05, 0.95), 
        legend.justification = c("left", "top"),
        axis.line = element_line(),
        axis.ticks = element_line(colour = "black"),
        axis.text.y = element_text(colour = "black", size = 9),
        axis.text.x = element_text(colour = "black", size = 12, angle = 45, hjust = 1),
        axis.title = element_text(colour = "black", size = 12))
p4
```

## Heatmap

Finally, we can visualize the data points in form of a heatmap.
I personally like to use `pheatmap` but the `ComplexHeatmap` package is also very nice:

* [pheatmap](https://cran.r-project.org/web/packages/pheatmap/pheatmap.pdf)
* [ComplexHeatmap](https://jokergoo.github.io/ComplexHeatmap-reference/book/)

```{r}
library(pheatmap)
cur_dat <- t(iris[1:4])
colnames(cur_dat) <- rownames(iris)
p5 <- pheatmap(cur_dat, scale = "row", 
         color = colorRampPalette(c("dark blue", "white", "dark red"))(100),
         annotation_col = data.frame(row.names = colnames(cur_dat),
                                     Species = iris$Species),
         show_colnames = FALSE,
         annotation_colors = list(Species = colour_vec),
         labels_row = sub("\\.", " ", rownames(cur_dat)))
```

## Stitiching it together

In the last part, I will show you how to put everything together using the `cowplot` or `patchwork` package:

* [cowplot](https://wilkelab.org/cowplot/index.html)
* [patchwork](https://patchwork.data-imaginist.com/articles/patchwork.html)

First, I will show you how `cowplot` works.
I will directly create nested `plot_grid` objects.

```{r cowplot}
library(cowplot)

upper_row <- p1
middle_row <- plot_grid(p2 + theme(legend.title = element_text(size = 10),
                                   legend.text = element_text(size = 8),
                                   legend.key.size = unit(0.1, "cm"),
                                   legend.position = c(1, 1.3)), 
                        p3 + theme(legend.title = element_text(size = 10),
                                   legend.text = element_text(size = 8),
                                   legend.key.width = unit(0.3, "cm")), 
                        p4 + theme(legend.title = element_text(size = 8),
                                   legend.text = element_text(size = 6),
                                   legend.position = c(0.05, 1.1),
                                   legend.key.size = unit(0.3, "cm")), 
                        ncol = 3, labels = c("B", "C", "D"))
bottom_row <- p5$gtable

final <- plot_grid(upper_row, middle_row, bottom_row, nrow = 3, labels = c("A", "", "E"))

ggsave(filename = "docs/final_figures/Fig_1.pdf", plot = final, width = 210, height = 290, units = "mm")
```

I have tried assembling the same figure using `patchwork` but didn't succeed :(
However, check it out at: [patchwork](https://patchwork.data-imaginist.com/articles/patchwork.html)
