---
title: "Session 1: R basic operations and data formats"
author: "Nils"
date: "`r Sys.Date()`"
output: html_notebook
---

This R notebook based in the markdown and R languages contains example code for the first session of the R course.
Here, we will learn how to perform basic operations in R, get to know the different data formats and play around with the [stats](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/00Index.html) package.

# Variables and the assignment operator

First, we will learn how to save a number (e.g. `42`) in a variable (e.g. `a`).
For this, we will use the assignment operator `<-` (and not `=`).

```{r example-variable}
a <- 42
a
```

The reason we don't use `=` is because it's reserved for setting parameters within a function call:

```{r example-assignment}
a <- c(1, 2, 3, 42)
var(x = a)
```

However, `var(x <- a)` will also work, but it will generate an object `x`, which is equal to `a`.

# Data classes

In R there are three basic data classes: `numeric`, `character` and `logical`.

## `numeric`

A number is of type `numeric`:

```{r numeric}
a <- 42
class(a)
is.numeric(a)
```

There are several different operations that can be performed on numeric values:

```{r numeric-examples}
# Addition
a <- 1 + 1
a

# Division
a <- 10/2
a

# Substraction
a <- 10 - 2
a

# Multiplication
a <- 10 * 2
a
```

## `character`

A string (e.g. a word) is of type `character`:

```{r character}
a <- "Hello"
class(a)
is.character(a)
```

Also a number of operations can be performed on characters:

```{r character-examples}
# Concatenate 1
a <- paste("Hello", "world!")
a

# Concatenate 2
a <- paste0("Hello", "world!")
a

# Splitting 1
a <- strsplit("Hello world", split = " ")
a

# Splitting 2
a <- strsplit(c("Hello world", "Test split"), split = " ")
a

# Grep
a <- grep(pattern = "b", x = c("b", "bb", "a"))
a

a <- grep(pattern = "b", x = c("b", "bb", "a"), value = TRUE)
a

a <- grepl(pattern = "b", x = c("b", "bb", "a"))
a

# Substitute
a <- sub(pattern = "b", replacement = "a", x = c("bba", "b", "c"))
a

a <- gsub(pattern = "b", replacement = "a", x = c("bba", "b", "c"))
a

# Quite useful
a <- gsub(pattern = "_[0-9]", replacement = "", x = c("marker1_1", "marker2_2", "marker3_3"))
a
```

## `logical`

Logicals are either `TRUE` or `FALSE`:

```{r logical}
a <- TRUE
class(a)
is.logical(a)
```

They are commonly used when checking expressions or comparing objects.

```{r logical-examples}
a <- 42
b <- 15
c <- a > b
c

c <- a < b
c
```

`TRUE` and `FALSE` can be abbreviated with `T` and `F`.
However, this is dangerous since `T` and `F` can be overwitten:

```{r T-and-F}
T
T <- 42
T
```

Functions usually list what data types they operate on. 
Have a lookt at the `sum` function by typing `?sum`.
You will see that the function takes either numeric, complex or logical vectors.

```{r sum-function}
sum(c(1, 2, 3, 4))
sum(c(TRUE, FALSE, TRUE))
```

## Coercion between data classes

Coercion means the conversion of a variable from one data class to another:

```{r coercion}
a <- 42
as.character(a)

a <- "Hello"
as.numeric(a)

a <- TRUE
as.numeric(a)

a <- 1
as.logical(a)
```

# Data containers

## `vector`

## `matrix`

## `list`

## `data.frame`

## `factor`

# Loops and the `if` statement

## `for` loop

## `apply`

## `if`

| || & &&

## `ifelse`

# The `stats` package

## Basic statistics

## "Advanced" statistics 
