---
title: "Session 1: R basic operations and data formats"
author: "Nils"
date: "`r Sys.Date()`"
output: html_notebook
---

This R notebook based in the markdown and R languages contains example code for the first session of the R course.
Here, we will learn how to perform basic operations in R, get to know the different data formats and play around with the [stats](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/00Index.html) package.

# Variables and the assignment operator

First, we will learn how to save a number (e.g. `42`) in a variable (e.g. `a`).
For this, we will use the assignment operator `<-` (and not `=`).

```{r example-variable}
a <- 42
a
```

The reason we don't use `=` is because it's reserved for setting parameters within a function call:

```{r example-assignment}
a <- c(1, 2, 3, 42)
var(x = a)
```

However, `var(x <- a)` will also work, but it will generate an object `x`, which is equal to `a`.

# Data classes

In R there are three basic data classes: `numeric`, `character` and `logical`.

## `numeric`

A number is of type `numeric`:

```{r numeric}
a <- 42
class(a)
is.numeric(a)
```

There are several different operations that can be performed on numeric values:

```{r numeric-examples}
# Addition
a <- 1 + 1
a

# Division
a <- 10/2
a

# Substraction
a <- 10 - 2
a

# Multiplication
a <- 10 * 2
a
```

## `character`

A string (e.g. a word) is of type `character`:

```{r character}
a <- "Hello"
class(a)
is.character(a)
```

Also a number of operations can be performed on characters:

```{r character-examples}
# Concatenate 1
a <- paste("Hello", "world!")
a

# Concatenate 2
a <- paste0("Hello", "world!")
a

# Splitting 1
a <- strsplit("Hello world", split = " ")
a

# Splitting 2
a <- strsplit(c("Hello world", "Test split"), split = " ")
a

# Grep
a <- grep(pattern = "b", x = c("b", "bb", "a"))
a

a <- grep(pattern = "b", x = c("b", "bb", "a"), value = TRUE)
a

a <- grepl(pattern = "b", x = c("b", "bb", "a"))
a

# Substitute
a <- sub(pattern = "b", replacement = "a", x = c("bba", "b", "c"))
a

a <- gsub(pattern = "b", replacement = "a", x = c("bba", "b", "c"))
a

# Quite useful
a <- gsub(pattern = "_[0-9]", replacement = "", x = c("marker1_1", "marker2_2", "marker3_3"))
a
```

## `logical`

Logicals are either `TRUE` or `FALSE`:

```{r logical}
a <- TRUE
class(a)
is.logical(a)
```

They are commonly used when checking expressions or comparing objects.

```{r logical-examples}
a <- 42
b <- 15
c <- a > b
c

c <- a < b
c
```

`TRUE` and `FALSE` can be abbreviated with `T` and `F`.
However, this is dangerous since `T` and `F` can be overwitten:

```{r T-and-F}
T
T <- 42
T
```

Functions usually list what data types they operate on. 
Have a lookt at the `sum` function by typing `?sum`.
You will see that the function takes either numeric, complex or logical vectors.

```{r sum-function}
sum(c(1, 2, 3, 4))
sum(c(TRUE, FALSE, TRUE))
```

## Coercion between data classes

Coercion means the conversion of a variable from one data class to another:

```{r coercion}
a <- 42
as.character(a)

a <- "Hello"
as.numeric(a)

a <- TRUE
as.numeric(a)

a <- 1
as.logical(a)
```

# Data objects

Here, we will look at different ways of storing numeric, logical or character entries in objects.

## `vector`

Vectors are objects that contain multiple entries.
They can be created in different ways:

```{r vector}
a <- vector(mode = "numeric", length = 5)
a
length(a)
class(a)
is.vector(a)

a <- c(0, 0, 0, 0, 0)
a
is.vector(a)

a <- rep_len(0, length.out = 5)
is.vector(a)

a <- 1:5
a
is.vector(a)

names(a) <- paste0("entry", a)
a
```

A number of functions take vectors as inputs:

```{r vector-functions}
sum(a)
var(a)
sd(a)
```

Vector operations include:

```{r vector-operations}
a * 5
a + 5
a - 5
a/5
```

Vectors can be subsetted as follows:

```{r vector-subsetting}
b <- a[3]
b

b <- a[c(2,3,4)]
b

b <- a[a > 2]
b

a[3] <- 100
a
```

## `matrix`

A matrix is a two-dimensional object containing `nrow` number of rows and `ncol` number of columns.

```{r matrix}
a <- matrix(data = NA, nrow = 10, ncol = 10)
a
dim(a)
nrow(a)
ncol(a)
colnames(a) <- paste("column", as.character(1:ncol(a)), sep = "_")
rownames(a) <- paste("row", as.character(1:nrow(a)), sep = "_")
is.matrix(a)
a
```

Let's construct a matrix with numeric entries:

```{r matrix-example}
a <- matrix(data = runif(100), nrow = 10, ncol = 10)
a
```

We can again perform matrix operations:

```{r matrix-operations}
a + 1
a * 2
a - 2
a/10

# Create a new vector
b <- 1:10
a * b

a %*% b
```

There are a number of convenient functions for matrices:

```{r matrix-functions}
rowSums(a)
colSums(a)
sum(a)
```

Subsetting a matrix works as follows:

```{r matrix-subsetting}
a[2,4]
a[1:3,1:5]

a[1:3,1:5] <- matrix(runif(15) + 10, ncol = 5, nrow = 3)
a
```

## `list`

A `list` can be constructed via the `list()` function:

```{r list}
a <- list(A = "test", B = 42)
a
is.list(a)
names(a) <- c("C", "D")
```

A list is the most flexible data object:

```{r list-examples}
a <- list(A = c("test1", "test2"), B = 1:10, C = a)
a
```

Subsetting can be performed as follows:

```{r list-subsetting}
a$A
a[1]
a[[1]]

# Works
a[1:2]

# Doesn't work as expected
a[[1:2]]

is.list(a[1])
is.list(a[[1]])
```

We see that `[` returns a `list` while `[[` returns the entry.

## `data.frame`

A `data.frame` is similar to a `list` (and a `matrix`) but the number of entries in each slot must be the same:

```{r data.frame}
df <- data.frame(A = 1:3, B = c("test1", "test2", "test3"))
df
is.data.frame(df)
typeof(df)
dim(df)
length(df)
ncol(df)
nrow(df)
```

Subsetting works as follows:

```{r data.frame-subsetting}
df$A
df[1]
df[1,1] # compare this to a[1,1]
```

The `data.frame` is the best option when storing numeric and non-numeric entries side-by-side.

## `factor`

By default, characters that are passed into the `data.frame()` functions are converted into a factor.

```{r factor}
df$B
is.factor(df$B)
```

Let's create a factor from scratch:

```{r create-factor}
a <- factor(c("test1", "test2", "test3"))
a
levels(a)
```

In a factor `levels` stores all unique entries in the factor. 
They remain in place when sub-setting a factor:

```{r factor-subsetting}
a[1]
```

Factors will come in handy when using `ggplot2` later on.

# Loops and the `if` statement

## `for` loop

## `apply`

## `if`

| || & &&

## `ifelse`

# The `stats` package

## Basic statistics

## "Advanced" statistics 
