---
title: "Session 2: R Basics - looping, if else, functions, stats package, reading in data"
author: "Nils"
date: "`r Sys.Date()`"
output: html_notebook
---

We will start with a small addition to session 1.

# Numeric vs integers and the numerical error

By default, R does not generate integer values.
They can be generated in the form of `1L, 2L, 45L, ...`.

```{r integers}
# FALSE
is.integer(1)

# TRUE
is.integer(1L)

# TRUE
is.integer(1:10)
```

When displaying floats in R, the numerical error is suppressed:

```{r numerical-error}
a <- 0.3/3
a
```

But testing does not result in the expected outcome:

```{r error-testing}
a == 0.1
```

It is therefore saver to use the `all.equal` function:

```{r equal-function}
all.equal(a, 0.1)
```

# Loops and the `if` statement

In the next section, we will learn about loops and the `if` statement in R.

## `while` loop

The `while` loop performs an operation as long as a condition is `TRUE`:

```{r while}
i <- 0

while (i < 5) {
    print(i)
    i <- i + 1
}
```

From my experience, the `while` loop in R is not that widely used for data analysis or programmning.

## `for` loop

The `for` loop can be used to iteratively perform some task.
Here is the lazy way of writing a for loop.

```{r for}
for (i in 1:10) print(i)
```

However, it is safer to write it like this:

```{r for2}
for (i in 1:10) { 
  print(i)
}
```

We have to use the `print` statement since variables are not displayed within loops.
Let's write a loop with higher functionality.
Here, `rnorm(x)` draws `x` random numbers from a normal distribution. 

```{r for3}
u1 <- rnorm(30)
u1
usq <- vector(mode = "numeric", length = 10)

for (i in 1:10) {
  usq[i] <- u1[i] * u1[i]
}
usq
```

For loops in R are very inefficient.
Most of the time, simple operations can just be vectorized:

```{r vectorized}
usq1 <- u1[1:10]
usq1 <- usq1*usq1

all.equal(usq, usq1)
```

## `apply`

A better alternative for the `for` loop is `apply()` and all its derivatives (`sapply`, `vapply`, `mapply`, `tapply`).
The main functions that we will discuss here are `lapply` and `sapply` for lists and vectors and `apply` for matrices (and arrays).

`lapply` performs operations on each entry of a list:

```{r lapply}
a <- as.list(1:10)
a

# Doesn't work: a * 2
a <- lapply(a, function(i){i * 2})
a
```

`sapply` simplifies the output to a vector (same as `lapply` when setting ``):

```{r sapply}
a_1 <- sapply(a, function(i){i + 10}, simplify = TRUE) # default behaviour
a_1

a_2 <- sapply(a, function(i){i + 10}, simplify = FALSE) # returns a list
a_2
```

Here is another example:

```{r sapply2}
a <- list(A = 1:5, B = 6:10)
sapply(a, sum)
```

In more complicated cases, `sapply` simplifies the output to a matrix:

```{r sapply-2}
a <- list(A = 1:5, B = rnorm(10), logi = c(TRUE, TRUE, FALSE, FALSE, FALSE))
sapply(a, quantile)
```

It is usually safer to use the `vapply` function that requires an expected output:

```{r vapply}
a <- list(name = "Nils", DOB = as.POSIXct("1988-10-27"))
a

sapply(X = a, FUN = class)

# Error
vapply(X = a, FUN = class, FUN.VALUE = character(1))
```

The `apply` function can be applied to matrices.
Here, `MARGIN` means the dimension to which the function is applied to (`1` meaning rows and `2` meaning columns).

```{r apply}
a <- matrix(data = runif(100), ncol = 10, nrow = 10)
row.sum <- apply(X = a, MARGIN = 1, FUN = sum)
col.sum <- apply(X = a, MARGIN = 2, FUN = sum)

# Same result as rowSums and colSums
all.equal(row.sum, rowSums(a))
all.equal(col.sum, colSums(a))
```

## `if`

The `if` statement is used for decision making:

```{r if}
a <- 42
if (a > 0) {
  print(paste(a, "is a positive number")) 
}
```

The `if` statement can be followed with an `else` option:

```{r if-else}
a <- -42
if (a > 0) {
  print(paste(a, "is a positive number")) 
} else {
  print(paste(a, "is a negative number or 0")) 
}
```

It is also possible to assess multiple statements at once.
For this, we can use `|` (or better `||`) or `&` (or better `&&`).

```{r logicals}
a <- runif(100)

# Finds all entries to a between 0.5 AND 0.7
a[(a > 0.5) & (a < 0.7)]

# This only tests the first argument
(a > 0.5) && (a < 0.7)

# Finds all entries smaller than 0.2 OR larger than 0.7
a[(a < 0.2) | (a > 0.7)]

# Tests only the first argument
(a < 0.2) || (a > 0.7)
```

Now let's use it in the `if` statement.
In `if` statements, it's better to use `||` instead of `|`.

```{r if-logical}
a <- "Hello"

if (length(a) == 1L && is.character(a)) {
  print(paste(expression(a), "contains a character of length 1"))
} else {
  print(paste(expression(a), "does not contain a character of length 1"))
}
```

## `ifelse`

To make decisions based on multiple checks, we can use the `ifelse` logic:

```{r ifelse}
a <- runif(100)
ifelse((a > 0.5) & (a < 0.7), "Between 0.5 and 0.7", "Other")
```

# Functions

# The `stats` package

The `stats` package contains most relvant statistical functions in R.
It is included in the base installation in R and does not need to be loaded.
We already used some `stats` functions above: `runif` and `rnorm`

We can use the stats package to generate random numbers:

```{r random-numbers}
# Uniform distribution
runif(10)

# Normal distribution
rnorm(10)

# Poisson distribution
rpois(10, lambda = 2)
```

The package also provides a function for correlation testing:

```{r correlation}
# Set seed to get reproducible results
set.seed(42)

# Generate correlated data
library(MASS)
test.cor <- MASS::mvrnorm(10, mu = c(0, 0),
                          Sigma = matrix(c(1, 0.67, 0.67, 1), nrow=2),
                          empirical = TRUE)

stats::cor(test.cor[,1], test.cor[,2])

# Default method is pearson
stats::cor.test(test.cor[,1], test.cor[,2])
stats::cor.test(test.cor[,1], test.cor[,2], method = "spearman")
```

The stats package also provides a function to perform T tests and Wilcoxon rank tests:

```{r tests}
test.cor <- MASS::mvrnorm(10, mu = c(0, 2),
                          Sigma = matrix(c(1, 0.67, 0.67, 1), nrow=2),
                          empirical = TRUE)

t.test(test.cor[,1], test.cor[,2])
wilcox.test(test.cor[,1], test.cor[,2])
wilcox.test(test.cor[,1], test.cor[,2], alternative = "less")
```
